---
title: 'Math/Stats Application Workshop: R Practice Set'
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: false
    theme: readable
    highlight: tango
    code_folding: show
    fig_width: 8
    fig_height: 5
    df_print: default
---

```{r setup, include=FALSE}
# This chunk configures every later chunk. `include=FALSE` means the chunk runs
# but neither its code nor its output appears in the knitted document.
#
# echo = TRUE      : show the R code in the output, not just the results
# fig.align        : center every figure
# dpi              : resolution of the generated PNG figures
#
# Warnings and errors are deliberately left visible. If a chunk complains, you
# should see the complaint.
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", dpi = 96)
```


# Math/Stats Application Workshop: R Practice Set

Sections:

1. Data Import
2. One Row per Country
3. 1-Tail Hypothesis Test: Do countries have more than five climate policies?
4. 2-Tail Means Comparison: Do high-income countries pass more climate laws?
5. Proportion Comparison Test: Are carbon-market countries more likely to have signed the Copenhagen Accord?
6. 2-Tail Means Comparison with Outliers: Do oil-rich countries pledge smaller reductions?
7. Group Comparisons: Does urbanization differ by income group?
8. Regression: Does national wealth predict the number of climate policies?


## How to use this practice set

Six new hypotheses on the same `policy_analysis.csv` you worked through in
`r_applications.Rmd`. Nothing here needs a technique you have not already
seen. Every test, every plot and every cleaning step has a worked example in
that file. When you get stuck, open the section listed below and adapt the
code.

| Practice section | Technique | Worked example in r_applications.Rmd |
|---|---|---|
| 3 | One-sample t-test, one-tailed: `t.test(x, mu, alternative)` | Section 3 |
| 4 | Two-sample Welch t-test: `t.test(x, y)` | Section 6 |
| 5 | Two-proportion test: `prop.test(x, n, correct = FALSE)`, then `sqrt()` for z | Section 5 |
| 6 | Welch t-test with an outlier check, then `wilcox.test(x, y)` | Section 6, including its index scatter and boxplot |
| 7 | One-way ANOVA: `aov()`, `summary()`, `TukeyHSD()` | Section 7 |
| 8 | Regression: `lm()`, `summary()`, `coeftest()` with `vcovHC(type = "HC3")` | Section 8 |

The data preparation moves you will need also have worked examples:
`select()` then `drop_na()` (every section of the reference), `group_by()`
with `filter(year == max(year))` (reference section 7), `group_by()` with
`summarise(mean(...))` (reference section 4), and `quantile()` (base R,
`?quantile`).

One function is new: `wilcox.test()`. It takes the same `x` and `y` arguments
as `t.test()` and compares the two groups by rank rather than by mean, so an
extreme value counts as "the largest" and nothing more. Run `?wilcox.test`
when you reach section 6.

Rules for every section:

- Fill in each empty chunk so that it does what the comment at the top of the
  chunk says.
- Write your interpretation in the "Your interpretation" block. Say what the
  statistic measures, what the p-value means under H0, whether you reject H0 at
  the 5% level, and what the result says about the economics. State the
  numbers.
- Sections 1 and 2 are done for you. Everything from section 3 on is yours.


```{r}
# package import
# load tidyverse library and here packages
pacman::p_load(tidyverse, here, lmtest, sandwich)
```

Same four packages as the reference file. `tidyverse` for data handling and
plots, `here` for paths, `lmtest` and `sandwich` for the robust standard
errors in section 8.




# 1. Data Import

```{r}
# import data

# here() builds an absolute path from the project root. It finds that root by
# looking for a marker file: the r_applications.Rproj sitting next to this .Rmd
# is one such marker. That works when this folder is the open project.
#
# If R was started somewhere else (say the top of the Git repository), here()
# locks onto that other root and the CSV will not be there. The second line is
# the fallback: when knitting, knitr sets the working directory to the folder
# holding this .Rmd, so the bare filename always resolves.
csv_path <- here("policy_analysis.csv")
if (!file.exists(csv_path)) csv_path <- "policy_analysis.csv"

# show_col_types = FALSE suppresses the block of text readr otherwise prints
# describing how it guessed each column's type.
policy_df <- read_csv(csv_path, show_col_types = FALSE)

# print data head
head(policy_df)
```

Same file as the reference: 10388 country-year rows, 47 columns, 212
countries, 1970 to 2018. The columns this practice set uses:

- `cum_policies`: cumulative count of national climate policies adopted up to
  that year.
- `cum_laws`: cumulative count of climate laws passed by the legislature.
- `cum_ETS`: cumulative count of emissions trading schemes in place.
- `partytopledge2`: 1 if the country registered a target under the 2009
  Copenhagen Accord, 0 if not.
- `rel_reduction_low3`: the low-bound emission reduction pledged under the
  Paris Agreement, in percent. Positive is a pledged cut, negative is a
  pledged increase.
- `oil_rents`: oil rents as a percent of GDP.
- `urban_pop`: share of the population living in urban areas, in percent.
- `GDP_percapita`, `Incomegroup`: as in the reference.




# 2. One Row per Country

All six hypotheses treat the country as the unit of observation, and all six
use each country's most recent year. Section 7 of `r_applications.Rmd` built
exactly this data frame with a grouped filter: `year == max(year)` evaluated
inside `group_by(country)` keeps the latest row for each country. Do it once
here and reuse `latest_df` in every section below.

```{r}
# one row per country: the most recent year each country appears in
latest_df <- policy_df %>%
  group_by(country) %>%
  filter(year == max(year)) %>%
  ungroup()

# how many rows, and which years survived?
nrow(latest_df)
table(latest_df$year)
```

212 rows, every one from 2018. That uniformity matters: the `cum_` columns
are cumulative counts, so the 2018 row holds each country's whole policy
history. No averaging over years is needed for them and none should be done.

```{r}
# missing values in the columns this practice set uses, latest year only
latest_df %>%
  select(Incomegroup, cum_policies, cum_laws, cum_ETS, partytopledge2,
         rel_reduction_low3, urban_pop, GDP_percapita, oil_rents) %>%
  summarise(across(everything(), ~ sum(is.na(.x))))
```

Two cautions before you start.

The latest row is not complete. `Incomegroup` is missing for 14 countries,
`partytopledge2` for 28, `rel_reduction_low3` for 59, `urban_pop` for 16 and
`GDP_percapita` for 29. Each section selects the columns it needs and then
drops NA rows, as the reference file does, so each section works with a
different number of countries. Report that number every time.

`oil_rents` is NA in 2018 for every one of the 212 countries. The oil series
stops in 2017. Section 6 has to go back to the full `policy_df` panel for
that variable.




# 3. 1-Tail Hypothesis Test: Do countries have more than five climate policies?

The Paris Agreement asks every party to submit a nationally determined
contribution and to deliver it through domestic measures. Delivery needs
instruments: a carbon price, an efficiency standard, a renewable target, a
land-use rule. `cum_policies` is the cumulative count of national climate
policies a country has adopted up to that year, and in the 2018 row it is the
country's full portfolio.

A claim you will hear at climate conferences is that most countries now have
a substantial climate policy portfolio. Put a number on "substantial": five
policies. This section asks whether the average country, in its latest year,
holds more than five.

The test is one-tailed and the direction is "greater", because the claim is
directional and it is stated before looking at the data. Reference section 3
used `"less"` because 3 was a ceiling and no upper tail existed. Here there is
no ceiling, so the direction is a substantive choice, and you must live with
it whatever the data say. Reread the reference file's discussion of one-tailed
versus two-tailed tests before you run this one, and pay attention to what a
one-tailed p-value can and cannot tell you.

### Our Hypotheses

- H0: The mean number of cumulative climate policies per country is 5.
- H1: The mean number of cumulative climate policies per country is greater
  than 5.

### Steps:

1. From `latest_df`, select `country`, `year` and `cum_policies`. Drop rows
   with NA.
2. Report n, the mean, the median and the standard deviation of
   `cum_policies`. Tabulate the distinct values with `table()`.
3. Plot the distribution as a histogram with one bar per integer value and a
   vertical line at the null value of 5. Label the axes and give it a title.
4. Run `t.test()` with `mu = 5` and `alternative = "greater"`.
5. Extract and print the t-statistic and the p-value. Print the full test
   object as well, so the confidence interval is visible.
6. Only after you have recorded the result of step 5: compute the p-value the
   same test gives with `alternative = "less"` and with `"two.sided"`, and the
   two-sided confidence interval. This is a check on your understanding of
   the tail, not a second chance at the test.

```{r}
# Step 1: build hypothesis_1_df from latest_df with country, year and
# cum_policies; drop NA rows; show head() and the row count
```

```{r}
# Step 2: n, mean, median and sd of cum_policies, then table() of its values
```

```{r h1-plot}
# Step 3: histogram of cum_policies with binwidth = 1 and boundary = -0.5 so
# each bar sits on an integer; geom_vline at 5 for H0 and a second geom_vline
# at the sample mean; axis labels, title and subtitle
```

```{r}
# Step 4 and 5: one-sample t.test on cum_policies with mu = 5 and
# alternative = "greater"; cat() the $statistic and $p.value; then print the
# whole test object
```

```{r h1-other-tails}
# Step 6: p-values from alternative = "less" and alternative = "two.sided" on
# the same data, and the two-sided $conf.int
```

### Your interpretation

*Write here: what the sign and size of the t-statistic mean, what the p-value
means under H0, whether you reject H0 at the 5% level, and what the sample
mean and the confidence interval say about the average country's policy
portfolio. Compare the mean with the median. Then answer this: does a large
p-value in a "greater" test mean there is no effect? Look at where the sample
mean sits relative to 5 before you answer, and say what the "less" and
two-sided p-values from step 6 add.*




# 4. 2-Tail Means Comparison: Do high-income countries pass more climate laws?

The reference file found no income gradient in two places. Section 5 showed
that high-income and other countries join the Paris Agreement at the same
rate, 96% each. Section 6 found no significant difference in their pledged
reductions. Signing is cheap, and a pledge is a number on a form. Legislation
is neither.

`cum_laws` is the cumulative count of climate laws a country has passed:
acts approved by a legislature, as distinct from executive strategies and
ministerial plans. A law binds future governments and takes parliamentary
time and legal drafting capacity. If income matters for climate action, it
may show up here, in domestic follow-through, rather than in international
commitments.

The comparison is High income against everyone else, the same split as
reference sections 5 and 6. Two-tailed, because a case can be made either
way: rich countries have more legislative capacity, but many poorer
countries have passed framework climate laws as a condition of climate
finance.

### Our Hypotheses

- H0: High-income and non-high-income countries have the same mean number of
  cumulative climate laws.
- H1: The two means differ.

### Steps:

1. From `latest_df`, select `country`, `Incomegroup` and `cum_laws`. Drop rows
   with NA. Note how many countries survive and why it is fewer than 212.
2. Split into a high-income data frame and a not-high-income data frame with
   `filter()`.
3. For each group report n, mean, median and variance. Compare the two
   variances before testing.
4. Plot both groups as a boxplot with jittered points and the group means
   marked. Label the axes and give it a title.
5. Run the two-sample `t.test()` with `x` and `y`, two-sided. Leave
   `var.equal` at its default of `FALSE`, so R runs Welch's version.
6. Print the t-statistic, the p-value and the full test object.
7. For comparison only: rerun with `var.equal = TRUE` and note how t and p
   change. Then run `wilcox.test()` on the same two vectors and note its
   p-value.

```{r}
# Step 1: build hypothesis_2_df from latest_df with country, Incomegroup and
# cum_laws; drop NA rows; print the row count and the number of NA values in
# latest_df$Incomegroup
```

```{r}
# Step 2: filter into hypothesis_2_hi_df (Incomegroup == "High income") and
# hypothesis_2_not_hi_df (Incomegroup != "High income")
```

```{r}
# Step 3: for each group, cat() n, mean, median and var of cum_laws
```

```{r h2-plot}
# Step 4: bind_rows() the two groups with a group label; boxplot with
# outlier.shape = NA, geom_jitter with height = 0, stat_summary for the mean;
# axis labels, title and subtitle
```

```{r}
# Step 5 and 6: two-sample Welch t.test with x = high income and y = not high
# income, alternative = "two.sided"; cat() the statistic and p-value; print
# the full object
```

```{r h2-checks}
# Step 7: the same test with var.equal = TRUE (Student's t), and wilcox.test()
# on the same two vectors; print each p-value
```

### Your interpretation

*Write here: the two means and their difference, what the t-statistic
measures, what the p-value means, whether you reject H0 at 5%, and the
confidence interval for the difference. Explain why Welch's test is the right
choice given the two variances and why its degrees of freedom are not a whole
number. Compare the medians with the means and say what that tells you about
which countries drive the gap. Then say what the result means for the claim
that income shows up in domestic legislation rather than in international
pledges.*




# 5. Proportion Comparison Test: Are carbon-market countries more likely to have signed the Copenhagen Accord?

An emissions trading scheme (ETS) is a carbon market: the government caps
total emissions, issues permits up to that cap, and lets firms trade them.
Building one is a large institutional investment. `cum_ETS` counts the
trading schemes a country has in place. In 2018 it is 0 for most countries, 1
for a few dozen, and 2 for one.

The Copenhagen Accord of 2009 asked countries to register emission targets
for 2020. `partytopledge2` is 1 for countries that did. Reference section 3
showed that most countries signed one or two of the three agreements rather
than all three, and Copenhagen is the one most often skipped.

The question is whether a country that has built a carbon market is more
likely to have registered a Copenhagen target. Both are commitments to a
quantified cap, so the two decisions should travel together.

### Our Hypotheses

- H0: The proportion that signed the Copenhagen Accord is the same for
  countries with an ETS and countries without one.
- H1: The two proportions differ.

### Steps:

1. From `latest_df`, select `country`, `cum_ETS` and `partytopledge2`. Drop
   rows with NA. Report the count and tabulate `cum_ETS`.
2. Create a logical column `has_ets` that is `TRUE` when `cum_ETS > 0`.
3. For each value of `has_ets`, count the countries, count the signatories,
   and compute the proportion that signed. The mean of a 0/1 column is the
   proportion.
4. Plot the two proportions as bars on a 0 to 1 axis, with the counts written
   on the bars. Label the axes and give it a title.
5. Run `prop.test()` with the counts read out of the data, not typed in. Use
   `correct = FALSE` to match the z-test in the Python version.
6. Take the square root of the chi-squared statistic to get the z-score.
   Print the chi-squared, the p-value and z.
7. List the countries that break the pattern: ETS without a Copenhagen target,
   and a Copenhagen target without an ETS.

Recall from reference section 5 that `prop.test()` reports a chi-squared
statistic on 1 degree of freedom. For two groups, z = sqrt(chi-squared). The
square root loses the sign, so read the direction from the two proportions.

```{r}
# Step 1 and 2: build hypothesis_3_df from latest_df with country, cum_ETS and
# partytopledge2; drop NA rows; mutate has_ets = cum_ETS > 0; print the row
# count and table(cum_ETS)
```

```{r}
# Step 3: group_by(has_ets) and summarise total = n(), signed =
# sum(partytopledge2), proportion = mean(partytopledge2); print it
```

```{r h3-plot}
# Step 4: geom_col of the two proportions with geom_text labels showing
# "signed of total (percent)"; coord_cartesian(ylim = c(0, 1.08)); a dashed
# geom_hline at the pooled proportion; axis labels, title and subtitle
```

```{r}
# Step 5 and 6: prop.test with x = c(signed with ETS, signed without) and
# n = c(total with ETS, total without), alternative = "two.sided",
# correct = FALSE; print the object; cat() chi-squared, p-value and
# sqrt(statistic) as z
```

```{r h3-exceptions}
# Step 7: filter() for has_ets & partytopledge2 == 0, then for !has_ets &
# partytopledge2 == 1; print both
```

### Your interpretation

*Write here: the two proportions and their difference in percentage points,
what the z-score measures, what the p-value means, whether you reject H0, and
what the confidence interval for the difference says. Then address two
cautions: how much does a test add when the gap is this large, and what does
the size of the smaller group do to the precision of its proportion? Finally,
look at the exceptions from step 7 and say whether this is evidence that
carbon markets cause countries to pledge, or evidence of something else.*




# 6. 2-Tail Means Comparison with Outliers: Do oil-rich countries pledge smaller reductions?

`oil_rents` is oil rents as a percent of GDP: the value of crude oil
production minus the cost of producing it, relative to the size of the
economy. An economy that earns a large share of its income from oil has more
to lose from global decarbonization, and its government may pledge less in
Paris as a result. That is the hypothesis.

Two data problems stand between the question and the test.

First, `oil_rents` is NA in the 2018 row for every country. The oil series
ends in 2017. So the oil measure has to be built from the full panel: average
each country's `oil_rents` over all the years it reports, then join that one
number per country to the latest-year pledge. Reference section 4 averaged
`GHG_percapita` over years in exactly this way.

Second, the outcome is `rel_reduction_low3`, the same variable as reference
section 6. You saw what Kiribati, Latvia and Madagascar did to that section.
They are still in the data. Check for outliers before testing, show the naive
test and the trimmed test side by side, and then run the rank-based
alternative that does not care about them.

Define oil-rich as the top quartile of mean oil rents among the countries
that have both a pledge and an oil series. That is: compute the 75th
percentile after the join, not before it.

Sign convention, from the reference: a positive `rel_reduction_low3` is a
pledged cut and a negative value is a pledged increase.

### Our Hypotheses

- H0: Oil-rich and other countries pledge the same mean Paris reduction.
- H1: The two means differ.

### Steps:

1. From the full `policy_df`, select `country` and `oil_rents`, drop NA, then
   `group_by(country)` and `summarise()` the mean as `mean_oil_rents`. Count
   the countries and check which years the oil series covers.
2. From `latest_df`, select `country` and `rel_reduction_low3`, drop NA.
3. `inner_join()` the two data frames by `country`. Report how many countries
   have both values.
4. Compute the 75th percentile of `mean_oil_rents` with `quantile()` on the
   joined data. Create `oil_rich` as "Oil-rich" when `mean_oil_rents` is at
   or above that value and "Rest" otherwise. Tabulate it.
5. For each group report n, mean, median and standard deviation. Compare each
   mean with its median.
6. Check for outliers before testing. Sort by `rel_reduction_low3` with
   `arrange()` and look at both ends. Plot the values as an index scatter as
   in reference section 6, colored by group, with the most extreme point
   labeled.
7. Run the naive Welch `t.test()` on all countries.
8. Remove the outliers you identified, state the rule you used, and rerun the
   summary and the test.
9. Run `wilcox.test()` on the full data as the robust alternative. Run it on
   the trimmed data too.
10. Plot the trimmed data as a boxplot by group with jittered points and a
    dashed line at zero.
11. Sensitivity: find the oil-rich country whose `mean_oil_rents` is closest
    to the cutoff, and compute the oil-rich mean pledge without it.

```{r}
# Step 1: oil_avg_df from policy_df: select country and oil_rents, drop_na(),
# group_by(country), summarise mean_oil_rents = mean(oil_rents) and
# n_years = n(); print nrow() and the range of years with non-NA oil_rents
```

```{r}
# Step 2: pledge_df from latest_df: select country and rel_reduction_low3,
# drop_na(); print nrow()
```

```{r}
# Step 3: hypothesis_4_df <- inner_join(oil_avg_df, pledge_df, by = "country");
# print nrow()
```

```{r}
# Step 4: oil_cut <- quantile(hypothesis_4_df$mean_oil_rents, 0.75); print it;
# mutate oil_rich = if_else(mean_oil_rents >= oil_cut, "Oil-rich", "Rest");
# table(oil_rich)
```

```{r}
# Step 5: group_by(oil_rich) and summarise n, mean, median, sd of
# rel_reduction_low3
```

```{r}
# Step 6a: arrange(rel_reduction_low3) and head(6); arrange(desc(...)) and
# head(4)
```

```{r h4-index-plot}
# Step 6b: index scatter. Add idx = seq_len(n()) after arranging by country;
# geom_point colored by oil_rich; geom_text label only for the one point below
# -10000; axis labels saying the x axis is spacing only; title and subtitle
```

```{r}
# Step 7: split into hypothesis_4_rich_df and hypothesis_4_rest_df; Welch
# t.test with x = oil-rich, y = rest, alternative = "two.sided"; cat() the
# statistic and p-value; print the full object
```

```{r}
# Step 8: outliers <- c(...); hypothesis_4_trim_df <- filter(!country %in%
# outliers); group summary again; Welch t.test again on the trimmed groups;
# cat() and print
```

```{r}
# Step 9: wilcox.test(x = oil-rich, y = rest) on the full data and on the
# trimmed data; print both
```

```{r h4-box-plot}
# Step 10: boxplot of the trimmed data by oil_rich with geom_jitter and a
# dashed geom_hline at 0; axis labels, title and subtitle
```

```{r h4-sensitivity}
# Step 11: among the oil-rich rows, arrange(mean_oil_rents) and head(3); then
# mean of rel_reduction_low3 for the oil-rich group without that lowest
# country
```

### Your interpretation

*Write here: the naive means, why one of them is absurd and which single
country produces it, how the trimmed test differs from the naive one, and
what the Mann-Whitney test adds. Report the statistic and p-value for each
version, in a small table. State whether you reject H0 under any of them.
Then say what the exercise teaches about a t-test on data with one value 180
times larger than the next, and whether the top-quartile definition of
oil-rich is itself a researcher degree of freedom, using what step 11
shows.*




# 7. Group Comparisons: Does urbanization differ by income group?

Development economics treats urbanization as part of structural
transformation: labor leaves agriculture, moves to cities, and productivity
rises. `urban_pop` is the share of a country's population living in urban
areas, in percent. `Incomegroup` is the World Bank's four-way classification.

Why this matters for climate policy: cities concentrate emissions from
buildings and transport, and they are also where policy reach is greatest. A
country that is 80% urban and one that is 30% urban face different abatement
problems. Before using urbanization as a control anywhere, it is worth
knowing how tightly it tracks income.

Four groups, so the tool is the one-way ANOVA from reference section 7,
followed by Tukey's HSD to see which pairs differ. Reference section 7 had
seven regions with very unequal sizes and found F = 4.5 with region
explaining 12% of the variation. Expect something different here.

### Our Hypotheses

- H0: Mean urban population share is the same in all four income groups.
- H1: At least one income group's mean differs from at least one other.

### Steps:

1. From `latest_df`, select `country`, `Incomegroup` and `urban_pop`. Drop
   rows with NA. Report the count.
2. Summarise by `Incomegroup`: n, mean, median and sd. Arrange by the mean.
3. Boxplot with jittered points, groups ordered by median, means marked.
   Label the axes and give it a title.
4. Fit `aov(urban_pop ~ Incomegroup, data = ...)`. Print `summary()` and
   extract the F-statistic and the p-value.
5. Run `TukeyHSD()` on the fitted model and print it.
6. Plot the Tukey intervals as in reference section 7.

```{r}
# Step 1: hypothesis_5_df from latest_df with country, Incomegroup and
# urban_pop; drop_na(); print nrow()
```

```{r}
# Step 2: group_by(Incomegroup) and summarise n_countries, mean, median, sd of
# urban_pop; arrange(mean)
```

```{r h5-boxplot}
# Step 3: boxplot with x = reorder(Incomegroup, urban_pop, FUN = median),
# geom_jitter, stat_summary(fun = mean) as a red diamond, coord_flip(); axis
# labels, title and subtitle
```

```{r}
# Step 4: aov(urban_pop ~ Incomegroup, data = hypothesis_5_df); summary();
# extract F and p with [[1]]$F[1] and [[1]]$"Pr(>F)"[1]; cat() both; print
# the summary table
```

```{r}
# Step 5: TukeyHSD() on the fitted model; print it
```

```{r h5-tukey-plot}
# Step 6: convert the Tukey matrix to a data frame with rownames_to_column(),
# rename `p adj` to p_adj, flag significant pairs, and draw geom_linerange
# intervals with a dashed line at zero; axis labels, title and subtitle
```

### Your interpretation

*Write here: the four group means in order, what F measures and what its
value says relative to the reference file's F of 4.5, what the p-value means,
whether you reject H0, and what share of the variation in urbanization lies
between income groups (compute it from the Sum Sq column). Then read the
Tukey table: which pairs differ, which pair does not, and why that one pair
fails when the others succeed. Finish with what the monotone pattern means
for using urbanization as a control variable alongside income.*




# 8. Regression: Does national wealth predict the number of climate policies?

Reference section 8 asked whether wealth bends the emissions curve and found
that it does not, once the standard errors were made robust. This section
asks a simpler version of the same question on a different outcome. The
Kuznets logic says rich countries demand environmental quality and their
governments supply policy in response. If so, the number of climate policies
a country has adopted should rise with income.

The model is a straight line: `cum_policies` on `log10(GDP_percapita)`, one
observation per country, latest year. Log base 10 is used because GDP per
capita in this sample runs from about 200 dollars to about 110,000, nearly
three orders of magnitude, and a slope on `log10` reads as "policies per
tenfold increase in income", which is a natural unit for a variable with that
range. The reference used `log()`, base e. Both are fine; the coefficient
just has different units. Choose one and say which.

### Our Hypotheses

- H0: The slope on `log10(GDP_percapita)` is zero. Wealth does not predict
  policy count.
- H1: The slope is not zero.

### Steps:

1. From `latest_df`, select `country`, `cum_policies` and `GDP_percapita`.
   Drop NA rows. Report the count.
2. Add a column `log10_gdp` equal to `log10(GDP_percapita)`. Check its range.
3. Scatter plot of `cum_policies` against `log10_gdp` with a fitted straight
   line. Label the axes and give it a title.
4. Fit `lm(cum_policies ~ log10_gdp, data = ...)` and print `summary()`.
5. Re-test the coefficients with HC3 robust standard errors:
   `coeftest(model, vcov = vcovHC(model, type = "HC3"))`.
6. Compare the classical and robust standard errors on the slope, and print
   the classical confidence interval for the slope with `confint()`.

```{r}
# Step 1 and 2: regression_df from latest_df with country, cum_policies and
# GDP_percapita; drop_na(); mutate log10_gdp = log10(GDP_percapita); print
# nrow() and range(log10_gdp)
```

```{r h6-plot}
# Step 3: geom_point of cum_policies against log10_gdp; geom_smooth(method =
# "lm", formula = y ~ x, se = TRUE); axis labels, title and subtitle
```

```{r}
# Step 4: reg_1 <- lm(cum_policies ~ log10_gdp, data = regression_df);
# print(summary(reg_1))
```

```{r}
# Step 5: coeftest(reg_1, vcov = vcovHC(reg_1, type = "HC3"))
```

```{r}
# Step 6: cat() the classical SE from summary(reg_1)$coefficients[2, 2] and
# the HC3 SE from the coeftest object [2, 2]; confint(reg_1)
```

### Your interpretation

*Write here: the slope and what one unit of `log10_gdp` means in dollars, the
t-statistic and p-value on the slope, whether you reject H0, and what the
R-squared says about how much of the variation in policy counts the line
explains. Compare the classical and HC3 standard errors and say whether the
choice changes the conclusion. Then compare this result with the reference
file's Kuznets regression: what did the squared term do there, and what does
the slope do here? Finish with what a null result like this one does and
does not say about wealth and climate policy.*
